---
title: "Data exploration"
output: html_notebook
---

```{r setup}

source("setup.R")

source("gen_funcs.R")

```


### Range of forecast accuracy

```{r Scoring accuracy for Melbourne}

q4.acc.df() %>%
  group_by(seas) %>%
  mutate(acc.avg = median(tm.Q4.acc, na.rm = TRUE)) %>%
  group_by(tm, tm.res, seas) %>%
  mutate(games = n()) %>%
  ungroup() %>%
  filter(tm.res != "draw") %>%
  # glimpse()
  ggplot(aes(x = tm.res, y = tm.Q4.acc)) +
  geom_hline(aes(yintercept = acc.avg), colour = "grey") +
  geom_boxplot() +
  facet_grid(~ seas) +
  # geom_text(aes(label = games, y = 0.1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Scoring accuracy for all teams",
       subtitle = "Percentage of scoring shots that are goals. Grey line represents season\'s median",
       y = "Teams scoring accuracy",
       x = "")

```

There definitely is an association between accuracy and winning for Melbourne, which is expected.

How is the distribution of relative accuracy?

```{r Relative scoring accuracy}
q4.acc.df() %>%
  group_by(tm, tm.res, seas) %>%
  mutate(games = n()) %>%
  ungroup() %>%
  filter(tm.res != "draw") %>% # ,
         # tm == "Melbourne") %>%
  ggplot(aes(x = tm.res, y = tm.rel.acc)) +
  geom_hline(aes(yintercept = 0), colour = "grey") +
  geom_boxplot() +
  facet_grid(~ seas) +
  # geom_text(aes(label = games, y = -0.6)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Relative scoring accuracy for all teams",
       subtitle = "Percentage of scoring shots that are goals, relative to opponent",
       y = "Teams scoring accuracy rel to opponent",
       x = "")

```
Not surprisingly, there is a clear association between the relative scoring accuracy and whether a team wins the match.


Looking at a scatterplot of the share of the total scoring shots versus the score of the total score, there is a small number of games where teams had a smaller number of scoring shots but still won the game.

```{r scoring shots vs total scores}

q4.acc.df() %>%
  filter(seas == 2017) %>%
  ggplot((aes(x = tm.Q4.ratio.sh, y = tm.Q4.ratio.sc))) +
  geom_point(aes(colour = tm.res)) +
  # geom_abline(aes(intercept = 0, slope = 1), linetype = 2) +
  geom_vline(aes(xintercept = 0.5), linetype = 2) +
  geom_hline(aes(yintercept = 0.5), linetype = 2) +
  labs(title = "Ratio of scoring shots vs ratio of final scores in 2017",
       x = "share of total scoring shots for the game",
       y = "share of total scores for the game") +
  geom_text(aes(label = "more shots but lost", x = 0.9,
                y = 0.45), data = tibble(x = 1), vjust = "bottom", hjust = "right") +
  geom_text(aes(label = "more shots and won", x = 0.9,
                y = 0.55), data = tibble(x = 1), vjust = "top", hjust = "right") +
  geom_text(aes(label = "fewer shots and lost", x = 0.1,
                y = 0.45), data = tibble(x = 1), vjust = "bottom", hjust = "left") +
  geom_text(aes(label = "fewer shots but won", x = 0.1,
                y = 0.55), data = tibble(x = 1), vjust = "top", hjust = "left")

```

About 1 game in 9 sees a team win despite fewer scoring shots.

```{r games won with fewer scoring shots}

q4.acc.df() %>%
  filter(tm.Q4.ratio.sh < 0.5, tm.Q4.ratio.sc >= 0.5) %>%
  ggplot((aes(x = seas))) +
  geom_bar() +
  coord_flip() +
  scale_y_continuous("number of games", position = "right",
                     sec.axis = sec_axis(trans = (~. / 198), labels = scales::percent,
                                         name = "Share of all games")) +
  labs(title = "Games won by the team with fewer scoring shots",
       subtitle = "About 1 game in 9 sees a team win despite fewer scoring shots",
       x = "season")

```



### Distribution of winning margins

```{r}

calc.margin() %>%
  select(seas, tm, home, tm.Q4.lead.abs) %>%
  mutate(seas = factor(seas)) %>%
  ggplot(aes(x = tm.Q4.lead.abs)) +
  # geom_histogram(binwidth = 6) +
  # geom_freqpoly(binwidth = 6) +
  geom_density(aes(colour = seas)) +
  scale_y_continuous(labels = scales::percent) +
  # facet_grid(~ seas)
  labs(title = "Density plot of margins for both winners and losers",
       x = "Winning score vs. losing score")

```
This chart doesn't really tell me much, other than 2017 and 2014 tended to have more close games than the other seasons.

What about if we look at the distribution of margins.

```{r}

calc.margin() %>%
  filter(home == 1) %>%
  mutate(Q4.margin.abs = abs(tm.Q4.lead.abs)) %>%
  ggplot(aes(x = Q4.margin.abs)) +
  # geom_bar(binwidth = 12) +
  geom_histogram(aes(y = 12 * ..density..), binwidth = 12, colour = "#969696") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = seq(0, 150, 24), minor_breaks = seq(0, 150, 12)) +
  facet_wrap(~ seas) +
  labs(title = "Frequency of different margins",
       y = "frequency of margin",
       x = "points margin")

```

```{r Frequency of different margins absolute}
calc.margin() %>%
  filter(home == 1) %>%
  mutate(Q4.margin.abs = abs(tm.Q4.lead.abs),
         seas = factor(seas),
         home = factor(home, levels = c(0, 1), labels = c("away", "home"))) %>%
  ggplot(aes(x = Q4.margin.abs, colour = seas)) +
  stat_ecdf(geom = "step", pad = FALSE) +
  geom_hline(aes(yintercept = 0.5), colour = "grey") +
  labs(title = "Frequency of different margins",
       y = "Cumulative frequency of games",
       x = "Final margin in points") +
  coord_cartesian(xlim = c(0, 75)) +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.2))

```

```{r Frequency of different margins relative}
calc.margin() %>%
  filter(home == 1) %>%
  mutate(seas = factor(seas),
         home = factor(home, levels = c(0, 1), labels = c("away", "home"))) %>%
  ggplot(aes(x = tm.Q4.lead.rel, colour = seas)) +
  geom_vline(aes(xintercept = 0.5), colour = "grey") +
  stat_ecdf(geom = "step", pad = FALSE) +
  labs(title = "Frequency of home team's relative score",
       y = "Cumulative number of games",
       x = "Home team's share of the total score, i.e. >50% = win") +
  coord_cartesian(xlim = c(0.1, 0.9)) +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.2)) +
  scale_x_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.1)) +
  geom_hline(aes(yintercept = 0.5), colour = "grey") +
  facet_wrap( ~ home)

```


```{r Frequency of different margins relative home vs away}
calc.margin() %>%
  # filter(home == 1) %>%
  mutate(seas = factor(seas),
         home = factor(home, levels = c(0, 1), labels = c("away", "home"))) %>%
  ggplot(aes(x = tm.Q4.lead.rel, colour = home)) +
  geom_hline(aes(yintercept = 0.5), colour = "grey") +
  stat_ecdf(geom = "step", pad = FALSE) +
  labs(title = "Frequency of home teams relative score",
       y = "Cumulative number of games",
       x = "Home team's share of the total score, i.e. >50% = win") +
  coord_cartesian(xlim = c(0.1, 0.9)) +
  scale_y_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.2)) +
  scale_x_continuous(labels = scales::percent,
                     breaks = seq(0, 1, by = 0.2)) +
  geom_vline(aes(xintercept = 0.5), colour = "grey") +
  facet_wrap( ~ seas)

```