---
title: "Ratings Notebook"
output: html_notebook
---

**Objectives**

1. Develop function to build a tibble with each team's rating after each round.
    + Factors in the season changeovers.
    + Takes different measures into account, e.g. scoring shots and normalised margins.
2. Develop function to calculate the rating accuracy:
    + Win/loss accuracy.
    + Margin accuracy.
3. Optimise the parameters.
4. Develop data visualisation.


```{r setup}

source("setup.R")

source("gen_funcs.R", local = TRUE)

# Load data for analysis
# source("gen_usecases.R")

theta <- 400
k.fact <- 40
p.fact <- 0.0464  # taken from The Arc Footy

# Select game statistic to base ratings on
optim.by <- "tm.Q4.lead.abs" #also tm.Q4.lead.rel.sd, tm.Q4.lead.abs

```

```{r load.results}

afl.results <- read_rds(paste0("../", data, "/", "afl.results.rds"))

# vector of seasons in the results
seas.rng <- afl.results %>%
   pull(seas) %>%
   unique() %>%
   sort(.)

# vector of teams
teams <- afl.results %>%
  select(tm = tm1) %>%
  distinct() %>%
  arrange(tm) %>%
  mutate(rating = 1500)  # Use 1500 as the initial rating

```


## Calculate Ratings after each Round

```{r calc.ratings}

seas.ratings <- vector("list", length(seas.rng))

for (s in seq_along(seas.rng)) {
  
  seas.sel <- seas.rng[[s]]
  
  afl.res.seas <- calc.margin(afl.results) %>%
    filter(seas == seas.sel, home == 1)
    # We only need 1 record per game, i.e. from home team's perspective
  
  # vector of each round in the season
  seas.rnds <- afl.res.seas %>%
    pull(., rnd) %>%
    unique(.) %>%
    sort(.)
  
  # Set up object to store round by round ratings 
  rnd.ratings <- vector("list", length(seas.rnds))
  
  # Set up ratings for round 1
  if (s == 1) { init.ratings <- teams }
  rnd.ratings[[1]] <- init.ratings
  
  # Move through each played round and update team ratings
  for (r in seq_along(seas.rnds)) {
    # print(seas.rnds[[r]])
    rnd.ratings[[r + 1]] <- upd.ratings(afl.res.seas, seas.rnds[[r]], 
                                        rnd.ratings[[r]], optim.by)
  } 
  
  seas.ratings[[s]] <- bind_rows(rnd.ratings, .id = "rnd") %>%
    mutate(rnd = factor(rnd, levels = seas.rnds, labels = seas.rnds),
           rnd = as.integer(as.character(rnd)))
  
  # Set up last.ratings to initialise the next season, if required.
  init.ratings <- rnd.ratings[[r + 1]]
  
}

full.ratings <- bind_rows(seas.ratings, .id = "seas") %>%
  mutate(seas = factor(seas, labels = seas.rng),
         seas = as.integer(as.character(seas))) %>%
    arrange(tm, seas, rnd) %>%
    group_by(tm) %>%
    mutate(Q4.diff.pred = lead(Q4.diff.pred),
           win.prob = lead(win.prob)) %>%  # Predicted margin referred to previous round whereas rating refers to the start of that round
    ungroup()

```

## Calculate the rating accuracy

```{r calc.acc.res}

act.with.pred(afl.results, optim.by) %>%
  group_by(seas) %>%
  summarise(tip.sc = sum(pred.res, na.rm = TRUE),
            marg.diff.avg = mean(marg.diff.abs))

```

**Compare accuracy with betting agencies**

The tipping score is comparable and my model's margin accuracy only slightly off (without any tuning yet)!

Betting odds sourced from data extracted from http://www.aussportsbetting.com/data/historical-afl-results-and-odds-data/

* Data copied 4 March 2018

```{r bet.agencies}

aus.bet.raw <- read_excel(paste0("../", data, "/", "afl.xlsx"), sheet = "Data", col_names = TRUE, skip = 1)

aus.bet <- aus.bet.raw %>%
  left_join(tm.map, by = c(`Home Team` = "tm.alt1")) %>%
  left_join(tm.map, by = c(`Away Team` = "tm.alt1"), suffix = c(".hm", ".aw")) %>%
  select(date = Date,
         tm.hm,
         tm.aw,
         final = `Play Off Game?`,
         Q4.diff.pred = `Home Line Close`,
         opp.pred = `Away Line Close`,
         tm1.Q4.tot = `Home Score`,
         tm2.Q4.tot = `Away Score`) %>%
  mutate(date = as_date(date),
         seas = year(date),
         Q4.diff.pred = -Q4.diff.pred,
         opp.pred = -opp.pred,
         tm1.Q4.lead.abs = tm1.Q4.tot - tm2.Q4.tot,
         pred.res = if_else(sign(Q4.diff.pred) == sign(tm1.Q4.lead.abs), 1, 0))

aus.bet %>%
  filter(is.na(final),
         as.character(seas) >= "2013") %>%
  group_by(seas) %>%
  summarise(tip.sc = sum(pred.res, na.rm = TRUE),
            marg.diff.avg = mean(abs(Q4.diff.pred - tm1.Q4.lead.abs), na.rm = TRUE))

```

**Compare model with betting agency**

There are no obvious problems with the model predictions compared to the betting agency predictions, based on the games with different predictions and top 10 difference in margines.

Note that this code is messy and should be tied in due course!

```{r comp.model.bet}

act.with.pred(afl.results, optim.by)%>%
  filter(seas == 2017, home == 1) %>%
  left_join(aus.bet, by = c("seas", "tm" = "tm.hm", "game.date" = "date"),
            suffix = c(".mod", ".bet")) %>%
  select(seas:Q4.diff.pred.tm, rating.opp, pred.hm:marg.diff.abs,
         Q4.diff.pred.bet = Q4.diff.pred,
         pred.res.bet) %>%
  rename(hm.mod.pred = pred.hm,
         hm.act = act.hm,
         mod.res = pred.res.mod,
         mod.act.diff.abs = marg.diff.abs,
         bet.res = pred.res.bet) %>%
  mutate(bet.act.diff.abs = abs(Q4.diff.pred.bet - tm.Q4.lead.abs),
         bet.mod.act.diff = abs(Q4.diff.pred.bet - Q4.diff.pred.tm)) %>%
  filter(bet.res != mod.res) %>%
  group_by(mod.res) %>%
  arrange(mod.res, -bet.mod.act.diff) %>%
  top_n(10, bet.mod.act.diff) %>%
  select(mod.res,
         game.date,
         tm, opp,
         hm.rtng = rating.hm,
         op.rtng = rating.opp,
         home.marg = tm.Q4.lead.abs,
         mod.marg = Q4.diff.pred.tm,
         bet.marg = Q4.diff.pred.bet
         )


```


## Visualise the ratings

```{r fig.width=13, fig.height=6}

rtngs.plot.data <- calc.margin(afl.results) %>%
  select(seas:home, !!as.name(optim.by)) %>%
  filter(seas == 2017) %>%
  # mutate(seas = factor(seas, levels = seas.rng, labels = seas.rng),
  #        rnd = factor(rnd)) %>%
  left_join(full.ratings, by = c("seas", "rnd", "tm")) %>%
  rename(rating.tm = rating) %>%
  left_join(full.ratings, by = c("seas", "rnd", "opp" = "tm"), suffix = c(".tm", ".aw")) %>%
  rename(rating.opp = rating) %>%
  left_join(tm.map, by = c("opp" = "tm")) %>%
  mutate(opp.abbr = tm.abbr,
         lbl = paste0(opp.abbr, ": ", if_else(tm.Q4.lead.abs > 0, "+", ""), tm.Q4.lead.abs),
         tm.win = factor(sign(tm.Q4.lead.abs), levels = -1:1, labels = c("loss", "draw", "win")),
         op.win = factor(sign(-tm.Q4.lead.abs), levels = -1:1, labels = c("loss", "draw", "win")))

# rtngs.plot.data %>%
#   glimpse()

rtngs.plot.data %>%
  # glimpse()
  ggplot(aes(x = rnd, y = rating.tm)) +
  geom_hline(aes(yintercept = 1500), colour = "black", alpha = 0.5) +
  geom_line(aes(group = tm), colour = "grey", na.rm = TRUE) +
  geom_line(data = rtngs.plot.data %>% filter(tm == "Melbourne"),
            aes(y = rating.tm), na.rm = TRUE, colour = "blue") +
  geom_point(data = rtngs.plot.data %>% filter(tm == "Melbourne"),
            aes(y = rating.tm, fill = tm.win), na.rm = TRUE, shape = 21, colour = "blue") +
  geom_point(data = rtngs.plot.data %>% filter(tm == "Melbourne"),
            aes(y = rating.opp, fill = op.win), na.rm = TRUE, shape = 21, colour = "black") +
  geom_text(data = rtngs.plot.data %>% filter(tm == "Melbourne"),
             aes(label = lbl, y = rating.opp), 
            size = 3, angle = 90, colour = "black", hjust = "bottom",
            nudge_y = 5, alpha = 0.75) +
  labs(title = "Melbourne's rating through 2017",
       x = "round", y = "rating at start of round") +
  scale_fill_manual("", values = c("win" = "blue", "loss" = "white", "draw" = "grey")) +
  theme(legend.position = "bottom") +
  scale_x_continuous(limits = c(1, 23))

```

**View ratings ladder**

```{r ratings.ladder}
full.ratings %>%
  filter(seas == max(seas),
         !is.na(rnd)) %>%
  filter(rnd == max(rnd)) %>%
  ggplot(aes(x = fct_reorder(tm, rating), y = rating, 
             colour = tm, fill = tm)) +
  geom_col(show.legend = FALSE) +
  geom_hline(aes(yintercept = 1500), colour = "grey") +
  coord_flip(ylim = c(1250, 1750)) +
  labs(title = "Ratings ladder after last round",
       x = "")

```

**View faceted win/loss**  

```{r fig.width=13}

rtngs.plot.data <- calc.margin(afl.results) %>%
  select(seas:home, !!as.name(optim.by)) %>%
  filter(seas == 2017) %>%
  left_join(full.ratings, by = c("seas", "rnd", "tm")) %>%
  rename(rating.tm = rating) %>%
  left_join(full.ratings, by = c("seas", "rnd", "opp" = "tm"), suffix = c(".tm", ".aw")) %>%
  rename(rating.opp = rating) %>%
  left_join(tm.map, by = "tm") %>%
  left_join(aus.bet.merge(aus.bet), by = c("game.date" = "game.date", "tm" = "tm")) %>%
  mutate(mod.corr = factor(sign(tm.Q4.lead.abs * Q4.diff.pred.tm),
                           levels = -1:1,
                           labels = c("Inc", "Draw", "Corr")),
         bet.tip = sign(tm.Q4.lead.abs * bet.pred.marg),
         bet.tip = factor(2 - bet.tip, levels = 1:3, labels = c("correct", "draw", "incorrect")))


rtngs.plot.data %>%
  # filter(tm == "North Melbourne", rnd == 15) %>%
  select(tm.abbr, rnd,
         actual = tm.Q4.lead.abs,
         model = Q4.diff.pred.tm,
         bet = bet.pred.marg,
         mod.corr,
         bet.tip,
         rating.tm) %>%
  gather(key = "meas",
         value = "margin",
         actual:bet) %>%
  mutate(tm.abbr = fct_reorder2(tm.abbr, rnd, rating.tm),
         meas = factor(meas, levels = c("bet", "model", "actual"))) %>%
  ggplot(aes(x = meas, y = margin, fill = meas)) +
  scale_y_continuous(breaks = seq(-60, 60, 24)) +
  geom_rect(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf, 
            aes(fill = mod.corr, linetype = bet.tip), alpha = 0.25, colour = "black") +
  geom_hline(aes(yintercept = 0), colour = "black", alpha = 0.5) +
  geom_bar(stat = "identity") +
  scale_fill_manual("",
                      values = c("actual" = "#01665e", "model" = "#8c510a", 
                                 "bet" = "#d8b365", "Corr" = "#5ab4ac", "Inc" = "white", "Draw" = "grey"),
                      labels = c("actual" = "actual margin", "model" = "model\`s margin", 
                                 "bet" = "bet\'s margin", "Corr" = "model correct", "Inc" = "", "Draw" = "draw")) +
  coord_flip(ylim = c(-60, 60)) +
  labs(subtitle = "each tick = 4 goals", y = "",
       title = "2017 Round by Round Margins and Tipping Accuracy") +
  facet_grid(tm.abbr ~ rnd, switch = "y") +
  theme(axis.text.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        strip.text.y = element_text(size = 7, angle = 180),
        legend.position = "bottom",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7))


```


### Plot Round-by-Round Scores

**Round by Round Score**  

```{r calc.round.by.round}

rnd.by.rnd.res <- act.with.pred(afl.results, optim.by)%>%
  filter(seas == 2017, home == 1) %>%
  left_join(aus.bet, by = c("seas", "tm" = "tm.hm", "game.date" = "date"),
            suffix = c(".mod", ".bet")) %>%
  select(seas:Q4.diff.pred.tm, rating.opp, pred.hm:marg.diff.abs,
         Q4.diff.pred.bet = Q4.diff.pred,
         pred.res.bet) %>%
  rename(hm.mod.pred = pred.hm,
         hm.act = act.hm,
         mod.res = pred.res.mod,
         mod.act.diff.abs = marg.diff.abs,
         bet.res = pred.res.bet) %>%
  mutate(bet.act.diff.abs = abs(Q4.diff.pred.bet - tm.Q4.lead.abs)) %>%
  group_by(seas, rnd) %>%
  summarise(games = n(),
            mod.res.sc = sum(mod.res, na.rm = TRUE),
            bet.res.sc = sum(bet.res, na.rm = TRUE),
            mod.marg.mad = mean(mod.act.diff.abs, na.rm = TRUE),
            bet.marg.mad = mean(bet.act.diff.abs, na.rm = TRUE)) %>%
  group_by(seas) %>%
  arrange(rnd) %>%
  mutate(games.cum = cumsum(games),
         mod.res.sc.cum = cumsum(mod.res.sc),
         bet.res.sc.cum = cumsum(bet.res.sc),
         mod.marg.mad.avg = cumsum(mod.marg.mad * games) / games.cum,
         bet.marg.mad.avg = cumsum(bet.marg.mad * games) / games.cum) %>%
  ungroup()

rnd.by.rnd.res %>%
  gather(measure, score, mod.res.sc, bet.res.sc) %>%
  mutate(measure = fct_recode(measure, "model" = "mod.res.sc", 
                              "bet" = "bet.res.sc")) %>%
  ggplot(aes(x = rnd, y = score, fill = measure)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(breaks = 0:9) +
  scale_x_continuous(breaks = 1:23) +
  labs(title = "Round by round tipping score",
       x = "") +
  scale_fill_manual("",
                    values = c("bet" = "#d8b365", "model" = "#8c510a")) +
  theme(legend.position = "bottom")

```

**Cumulative Tipping Score**  

This isn't particularly useful!
```{r cum.tip.score}

rnd.by.rnd.res %>%
  ggplot(aes(x = rnd)) +
  geom_line(aes(y = bet.res.sc.cum, colour = "bet")) +
  geom_line(aes(y = mod.res.sc.cum, colour = "model")) +
  scale_colour_manual("",
                    values = c("bet" = "#d8b365", "model" = "#8c510a")) +
  scale_x_continuous(breaks = 1:23, minor_breaks = NULL) +
  labs(title = "Cumulative tipping score",
       x = "", y = "") +
  theme(legend.position = "bottom")

  

```

**Model vs Betting Agency**  

```{r cum.model.vs.bet}

rnd.by.rnd.res %>%
  mutate(bet.vs.mod = bet.res.sc.cum - mod.res.sc.cum) %>%
  ggplot(aes(x = rnd)) +
  geom_hline(aes(yintercept = 0), colour = "grey") +
  geom_line(aes(y = bet.vs.mod)) +
  geom_point(aes(y = bet.vs.mod)) +
  scale_x_continuous(breaks = 1:23, minor_breaks = NULL) +
  labs(title = "Cumulative betting agency vs model tipping score",
       x = "", y = "Betting agency score vs model")

  

```


**Model vs Betting Agency Margin**  

```{r margin.model.vs.bet}

rnd.by.rnd.res %>%
  gather(measure, score, bet.marg.mad.avg, mod.marg.mad.avg) %>%
  mutate(measure = fct_recode(measure, "model" = "mod.marg.mad.avg", 
                              "bet" = "bet.marg.mad.avg")) %>%
  # mutate(bet.vs.mod = bet.marg.mad.avg - mod.marg.mad.avg) %>%
  ggplot(aes(x = rnd, y = score, colour = measure)) +
  geom_line() +
  scale_x_continuous(breaks = 1:23, minor_breaks = NULL) +
  scale_colour_manual("",
                      values = c("bet" = "#d8b365", "model" = "#8c510a")) +
  labs(title = "Betting agency vs model margin forecast",
       subtitle = "Average absolute difference between forecast and actual margin - lower is better",
       x = "", y = "Difference between forecast and actual margin")

  

```

**Model vs Betting Agency Margin**  

```{r cum.marg.model.vs.bet}

rnd.by.rnd.res %>%
  mutate(bet.vs.mod = bet.marg.mad.avg - mod.marg.mad.avg) %>%
  ggplot(aes(x = rnd)) +
  geom_hline(aes(yintercept = 0), colour = "grey") +
  geom_line(aes(y = bet.vs.mod)) +
  scale_x_continuous(breaks = 1:23, minor_breaks = NULL) +
  labs(title = "Cumulative betting agency vs model margin forecast",
       subtitle = "Average absolute difference between forecast and actual margin - lower is better",
       x = "", y = "Betting agency score vs model")

  

```


#### Box plot of margins


```{r box.plot.margins, fig.width=10}

act.w.pred.bet <- act.with.pred(afl.results, optim.by)%>%
  filter(seas == 2017, home == 1) %>%
  left_join(aus.bet, by = c("seas", "tm" = "tm.hm", "game.date" = "date"),
            suffix = c(".mod", ".bet")) %>%
  select(seas:Q4.diff.pred.tm, rating.opp, pred.hm:marg.diff.abs,
         Q4.diff.pred.bet = Q4.diff.pred,
         pred.res.bet) %>%
  rename(hm.mod.pred = pred.hm,
         hm.act = act.hm,
         mod.res = pred.res.mod,
         mod.act.diff.abs = marg.diff.abs,
         bet.res = pred.res.bet) %>%
  mutate(bet.act.diff.abs = abs(Q4.diff.pred.bet - tm.Q4.lead.abs))

# act.w.pred.bet %>% glimpse()

act.w.pred.bet %>%
  select(seas, rnd, mod.act.diff.abs, bet.act.diff.abs) %>%
  gather(key = pred.src, value = pred.act.diff, mod.act.diff.abs, bet.act.diff.abs) %>%
  # mutate( rnd = factor(rnd),
  mutate( pred.src = fct_recode(pred.src, "model" = "mod.act.diff.abs", 
                              "bet" = "bet.act.diff.abs"),
          rnd.rng = factor((rnd - 1) %/% 5, levels = 0:4, labels = c("1-5", "6-10", "11-15", "16-20", "20-23"))) %>%
  group_by(seas, rnd.rng, pred.src) %>%
  mutate(pred.act.mad = mean(pred.act.diff, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = rnd.rng, y = pred.act.diff, colour = pred.src)) +
  geom_boxplot() +  # I attempted using cut_interval but the colour dimension got lost
  geom_point(aes(y = pred.act.mad), position = position_dodge(.8)) +
  scale_colour_manual("",
                      values = c("bet" = "#d8b365", "model" = "#8c510a")) +
  labs(title = "Distribution of predicted margin errors for season 2017",
       subtitle = "Lower scores are better. Average errors shown inside boxes",
       y = "Predicted margin error",
       x = "Rounds")

```



#### Histogram of proportion of wins by probability bin

```{r chart.prob.preds}

pred.bins <- act.with.pred(afl.results, optim.by) %>%
  filter(seas == 2017) %>%
  arrange(rnd) %>%
  select(seas, rnd, tm, opp, win.prob.tm, act.hm) %>%
  mutate(prob.bin = ((win.prob.tm - 0.025) %/% 0.05) * 0.05,
         win.tm = (act.hm > 0)) %>%
  group_by(seas, prob.bin) %>%
  summarise(win.avg = mean(win.tm, na.rm = TRUE),
            games = n()) %>%
  ungroup()

pred.bins %>%
  # glimpse()
  # mutate(prob.bin = factor(prob.bin)) %>%
  ggplot(aes(prob.bin, win.avg)) +
  geom_col(stat = "identity", width = 0.03) +
  geom_abline(slope = 1, intercept = 0) +
  geom_text(aes(label = games, y = -0.1), size = 3) +
  coord_cartesian(xlim = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 1, 0.05), expand = c(0, 0.0)) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0.05)) +
  labs(title = "Probability performance",
       x = "Estimated probability of the home team winning",
       y = "Share of games won") 

```

